---
title: "Melanoma_CellTyping_2d"
output: html_document
date: "2024-02-29"
---

2d: SCTransform on each sample independently (without library/log normalization, ScaleData, or FindVariableFeatures), integration with Harmony, clustering and cell typing on merged data 

Integration strategy: Perform (SCT) normalization independently on each sample and find integration features across samples using Seurat; then merge these normalized Seurat objects, set variable features manually to integration features, and finally calculate PCs on this merged object. Run Harmony. 

**Cell Typing/Annotation**


```{r}
library(Seurat)
library(SeuratWrappers)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(clustree)
library(gridExtra)
library(ggridges)
library(DoubletFinder)
library(tidyverse)
require(ggridges)
require(ggplot2)
library(readr)
```


Parameters
```{r}

n.pcs <- 10
res.used <- 0.9
seed.use <- 123

```


*Get reference data*
```{r}

cell.genes <- read_tsv("/gpfs/gsfs12/users/Sherlock_Lung/CalebHartman/scMelanoma_Results/Melanoma_cell_type_markers.txt")

# filter to Primary Type

genes_to_check <- cell.genes %>% filter(Type=="Primary") %>% pull(gene)
genes_to_check <- as.character(unique(genes_to_check))
genes_to_check <- genes_to_check[genes_to_check %in% rownames(seurat_phase@assays$SCT)]



```


*Visualize individual markers*


Dot plot
```{r}
pdf(paste("Mel_Norm_2d_",n.pcs,"pcs_res",res.used,"_seed",seed.use,"_primary_marker_DotPlot.pdf", sep=""),  width = 12, height = 20)

DotPlot(seurat_phase, features=genes_to_check) + coord_flip()

dev.off()


```

Filtering
```{r}

genes_to_check <- cell.genes %>% filter(Type=="Primary")  %>% filter(gene%in%rownames(seurat_phase@assays$SCT))

#seurat_phase <- JoinLayers(seurat_phase)

data_temp <- as.data.frame(as.matrix(GetAssayData(object=seurat_phase[["SCT"]], layer="data")))

umap.coor <- Embeddings(seurat_phase,reduction="umap")

ggplot.list.umap <- list()
ggplot.list.2 <- list()





```

BoxPlot
```{r}

for(i in 1:nrow(genes_to_check)){
  gene.name <- genes_to_check$gene[i]
  cell.type <- genes_to_check$cell[i]
  gene.exp <- data_temp[gene.name,]
  clusters <- seurat_phase@meta.data$seurat_clusters
  
  df.umap <- as.data.frame(cbind(umap.coor, as.data.frame(t(gene.exp)),clusters))
  colnames(df.umap) <- c(colnames(umap.coor),"gene.exp","clusters")
  
  ggplot.list.umap[[i]] <- ggplot(df.umap, aes(UMAP_1, UMAP_2)) +
    geom_point(aes(colour=gene.exp), size=0.7) +
    scale_colour_gradientn(colours=c("lightgray", "blue"),limits=quantile(df.umap$gene.exp,probs=c(0.05,0.95)),oob=scales::squish) +
    labs(title=paste(gene.name,"(",cell.type,")",sep=""))+
    theme(plot.title=element_text(hjust=0.5))


  ggplot.list.2[[i]] <-ggplot(df.umap) +
    geom_boxplot(aes(x=clusters,y=gene.exp))  + 
    ggtitle(paste(gene.name,"(",cell.type,")",sep="")) + ylab("gene expression (log)")+
    theme(plot.title=element_text(hjust=0.5))
}


n <- length(ggplot.list.umap)
ncol <- floor(sqrt(n))

pdf(paste("Mel_Norm_2d_",n.pcs,"pcs_res",res.used,"_seed",seed.use,"_Boxplot_primary_markers.pdf", sep=""), width = 20, height = 20)
do.call("grid.arrange", c(ggplot.list.2, ncol=ncol))
dev.off()





```

Feature Plot
```{r}

pdf(paste("Mel_Norm_2d_",n.pcs,"pcs_res",res.used,"_seed",seed.use,"_UMAP_primary_markers.pdf", sep=""),width=30,height=30)

FeaturePlot(seurat_phase, reduction="umap",features=genes_to_check$gene,ncol=floor(sqrt(nrow(genes_to_check))),min.cutoff="q10",max.cutoff="q90",pt.size=0.1)

dev.off()


```



*Visualize average marker expression*

```{r}

require(ggridges)
require(ggplot2)

cell.genes <- cell.genes[as.character(cell.genes$Type)=="Primary",]
cell.types  <- as.character(unique(cell.genes$cell))
umap.coor <- Embeddings(seurat_phase,reduction = "umap")
ggplot.list <- list()
ggplot.list.2 <- list()

#seurat_phase <- JoinLayers(seurat_phase)
data_temp <- as.data.frame(as.matrix(GetAssayData(object = seurat_phase[["SCT"]], layer = "data")))


for(i in 1:length(unique(cell.types)))

{
   genes <- as.character(cell.genes$gene[which(cell.genes$cell==cell.types[i])])
   genes <- intersect(genes,rownames(data_temp))
   gene.exp <- colMeans(as.matrix(data_temp[genes,]))[row.names(umap.coor)]
   clusters <- seurat_phase@meta.data$SCT_snn_res.0.9
  
   temp <- as.data.frame(cbind(umap.coor, as.data.frame(gene.exp), as.data.frame(clusters)))
  
   ggplot.list[[i]] <- ggplot(temp, aes(UMAP_1, UMAP_2)) + 
   geom_point(aes(colour = gene.exp),size=1) + 
   scale_colour_gradient(low = "grey95", high = "red") + 
   labs(title = cell.types[i], subtitle = paste(genes, collapse = ", "))
   
   # Boxplot per cluster
   
   ggplot.list.2[[i]] <- ggplot(temp, aes(x = clusters, y = gene.exp)) + 
   geom_boxplot() + 
   ggtitle(cell.types[i]) + ylab("Average gene expression (log)")
  
}

# Plot all 
require(grid)
require(gridExtra)
require(gbm)
n <- length(ggplot.list)
nCol <- floor(sqrt(n))




```

BoxPlot
```{r}


# Expression per cluster boxplots 
pdf(paste("Mel_Norm_2d_",n.pcs,"pcs_res",res.used,"_seed",seed.use,"_boxplots_with_avg_expression_of_cell_markers.pdf", sep=""),20,20)

do.call("grid.arrange", c(ggplot.list.2, ncol=nCol))

dev.off()


```

*General Cluster Annotation*

```{r}

 seurat_phase@meta.data$seurat_cluster_id <- seurat_phase@meta.data$seurat_clusters

 seurat_phase@meta.data$main_cluster_id <- seurat_phase@meta.data$seurat_clusters

 cluster.id <- sort(as.numeric(unique(as.character(seurat_phase@meta.data$seurat_cluster_id))))

 free_annotation <- c("melanocytes, fibroblasts", #0
                      "fibroblasts", #1
                      "fibroblasts", #2
                      "T cells", #3
                      "melanocytes, fibroblasts", #4
                      "keratinocytes", #5
                      "mixed", #6
                      "mixed", #7
                      "melanocytes, fibroblasts", #8
                      "keratinocytes", #9
                      "vascular endothelial, keratinocytes", #10
                      "macrophages, DCs, monocytes", #11
                      "fibroblasts, Schwann cells", #12
                      "mixed", #13
                      "keratinocytes", #14
                      "vascular endothelial, keratinocytes", #15
                      "keratinocytes, sweat gland", #16
                      "melanocytes, fibroblasts", #17
                      "fibroblasts", #18
                      "sweat gland", #19
                      "melanocytes, fibroblasts", #20
                      "keratinocytes", #21
                      "T cells, keratinocytes", #22
                      "vascular endothelial, lymphatic endothelial", #23
                      "melanocytes, fibroblasts", #24
                      "neuron", #25
                      "fibroblasts" #26
                      

 )



seurat_phase@meta.data$main_cluster_id <-   as.character(plyr::mapvalues(seurat_phase@meta.data$seurat_clusters, from=cluster.id, to=free_annotation))

view(table(seurat_phase@meta.data$main_cluster_id))

save(seurat_phase, file='Mel_Norm_2d_Annotated_Cluster.RData')


```

Plot UMAP
```{r}
seurat_phase <- RenameIdents(object = seurat_phase, 
                      "0" = "melanocytes, fibroblasts",
                      "1" = "fibroblasts", 
                      "2" = "fibroblasts", 
                      "3" = "T cells", 
                      "4" = "melanocytes, fibroblasts", 
                      "5" = "keratinocytes", 
                      "6" = "mixed", 
                      "7" = "mixed", 
                      "8" = "melanocytes, fibroblasts", 
                      "9" = "keratinocytes", 
                      "10" = "vascular endothelial, keratinocytes", 
                      "11" = "macrophages, DCs, monocytes", 
                      "12" = "fibroblasts, Schwann cells", 
                      "13" = "mixed", 
                      "14" = "keratinocytes", 
                      "15" = "vascular endothelial, keratinocytes", 
                      "16" = "keratinocytes, sweat gland", 
                      "17" = "melanocytes, fibroblasts", 
                      "18" = "fibroblasts", 
                      "19" = "sweat gland", 
                      "20" = "melanocytes, fibroblasts", 
                      "21" = "keratinocytes", 
                      "22" = "T cells, keratinocytes", 
                      "23" = "vascular endothelial, lymphatic endothelial", 
                      "24" = "melanocytes, fibroblasts", 
                      "25" = "neuron", 
                      "26" = "fibroblasts") 
                      


# Plot the UMAP
pdf(paste("Mel_Norm_2d_",n.pcs,"pcs_res",res.used,"_seed",seed.use,"_Annotated_UMAP.pdf", sep=""),20,20)

DimPlot(object = seurat_phase, 
        reduction = "umap", 
        label = TRUE,
        label.size = 3,
        repel = TRUE)

dev.off()

#End of Script
```

END of SCRIPT