---
title: "Melanoma_normalization_2d"
output: html_document
date: "2024-02-29"
---

2d: SCTransform on each sample independently (without library/log normalization, ScaleData, or FindVariableFeatures), integration with Harmony, clustering and cell typing on merged data 

Integration strategy: Perform (SCT) normalization independently on each sample and find integration features across samples using Seurat; then merge these normalized Seurat objects, set variable features manually to integration features, and finally calculate PCs on this merged object. Run Harmony.  


```{r}
library(Seurat)
library(tidyverse)
library(RCurl)
library(cowplot)
library(scater)
library(SingleR)
library(celldex)
library(pheatmap)
library(viridis)
library(scRNAseq)
library(ExperimentHub)
library(scuttle)
library(POMA)
library(dplyr)
library(RColorBrewer)
library(harmony)

```


Split seurat object by condition/sample to perform cell cycle scoring and SCT on all samples individually

```{r}
# load in filtered seurat

#load("/gpfs/gsfs12/users/Sherlock_Lung/CalebHartman/scMelanoma_Results/final_filtered_seurat.RData")

split_seurat <- SplitObject(filtered_seurat, split.by = "sample")


```


Run the following loop to perform sctransform on all samples

```{r}
for (i in 1:length(split_seurat)) 
{
    split_seurat[[i]] <- SCTransform(split_seurat[[i]], vst.flavor = "v2")

}
```


```{r}
# Find most variable features across samples to integrate
integ_features <- SelectIntegrationFeatures(object.list = split_seurat, nfeatures = 2000) 

# Merge normalized samples
merged_seurat <- merge(x = split_seurat[[1]],
		       y = split_seurat[2:length(split_seurat)],
		       merge.data = TRUE)

DefaultAssay(merged_seurat) <- "SCT"

# Manually set variable features of merged Seurat object
VariableFeatures(merged_seurat) <- integ_features

# Calculate PCs using manually set variable features
seurat_phase <- RunPCA(merged_seurat, assay = "SCT")
```


*Evaluate cell cycle effects*
```{r}

# Load cell cycle markers
load("/gpfs/gsfs12/users/Sherlock_Lung/CalebHartman/scMelanoma_Results/cycle.rda")

#seurat_phase <- JoinLayers(seurat_phase)

# Score cells for cell cycle
seurat_phase <- CellCycleScoring(seurat_phase, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

# View cell cycle scores and phases assigned to cells                                 
View(seurat_phase@meta.data)               

```


*PCA for cell cycle effects*

```{r}

pdf("Cell_Cycle_PCA.pdf", width = 10, height = 12)

#seurat_phase <- RunPCA(seurat_phase, assay = "SCT")

print(seurat_phase[['pca']], dims = 1:5, nfeatures = 5)
DimHeatmap(seurat_phase, dims = 1, cells = 500, balanced = TRUE)
DimPlot(seurat_phase, reduction = "pca", group.by = "Phase", split.by = "Phase")

dev.off()

```

*PCA for Mitochondrial expression*

```{r}
pdf("Mitochondrial_PCA.pdf", width = 10, height = 12)

# Check quartile values
summary(seurat_phase@meta.data$mitoRatio)

# Turn mitoRatio into categorical factor vector based on quartile values
seurat_phase@meta.data$mitoFr <- cut(seurat_phase@meta.data$mitoRatio, 
                   breaks=c(-Inf, 0.001220, 0.002492, 0.004652, Inf), 
                   labels=c("Low","Medium","Medium high", "High"))

# the 'breaks' numbers correspond to the 1st quartile, median, and 3rd quartile respectively 

# Plot the PCA colored by mitoFr
DimPlot(seurat_phase, reduction = "pca", group.by= "mitoFr", split.by = "mitoFr")

dev.off()
```


*Save seurat_phase*

```{r}

saveRDS(seurat_phase, "seurat_phase.rds")


```


*Harmony*
```{r}
seurat_phase <- RunHarmony(seurat_phase, 
				group.by.vars = "sample", 
				reduction = "pca", assay.use = "SCT", reduction.save = "harmony")




```


*Elbow plot to determine number of PCs to keep*
```{r}

# Printing out the most variable genes driving PCs
print(x = seurat_phase[["harmony"]], dims = 1:10, nfeatures = 5)

ElbowPlot(seurat_phase)

```

```{r}
seurat_phase <- RunUMAP(seurat_phase, reduction = "harmony", assay = "SCT", dims = 1:10)

```



*Clustering after Harmony integrating*

FindNeighbors
FindClusters


```{r}
n.pcs <- 10
seed.use <- 123
res.used <- c(0.1, 0.3, 0.5, 0.7, 0.9)

seurat_phase <- FindNeighbors(seurat_phase, reduction = 'harmony')
seurat_phase <- FindClusters(seurat_phase,resolution=res.used) 
                             


```


*Visualize clusters*
```{r}
# explore resolutions 
seurat_phase@meta.data %>% view()


```

Plot UMAP at different resolutions
```{r}

#seurat_phase <- RunUMAP(seurat_phase, reduction = "harmony", assay = "SCT", dims = 1:n.pcs)

#-----------------------------------------

Idents(object = seurat_phase) <- "SCT_snn_res.0.1"

DimPlot(seurat_phase, label=TRUE, label.size = 6, reduction = "umap") 

#-----------------------------------------

Idents(object = seurat_phase) <- "SCT_snn_res.0.5"

DimPlot(seurat_phase, label=TRUE, label.size = 6, reduction = "umap")

#-----------------------------------------

Idents(object = seurat_phase) <- "SCT_snn_res.0.9"

DimPlot(seurat_phase, label=TRUE, label.size = 6, reduction = "umap") 



```



Plot FeaturePlot to find uninteresting sources of variation

```{r}

pdf("Mel_Norm_2d_Variation.pdf", width=30, height=30)

metrics <- c("nUMI", "nGene", "mitoRatio", "S.Score", "G2M.Score")

FeaturePlot(seurat_phase, reduction="umap", features=metrics, min.cutoff="q10", max.cutoff="q90", pt.size=0.1, label = TRUE, order = TRUE)

dev.off()

```


Save final version of 'seurat_phase' before cell typing 
```{r}

#saveRDS(seurat_phase, "seurat_phase_final.rds")

save(seurat_phase, file = "seurat_phase_final.RData")


# End of Script 
```

END of Script

